---
title: "Dada2 tutorial"
output: html_notebook
---
```{r}
library("dada2")
```

```{r}
path <- "~/MiSeq_SOP" # CHANGE ME to the directory containing the fastq files after unzipping.
list.files(path)
```

```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

```{r}
plotQualityProfile(fnRs[1:4])
```

```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(240,160),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE
head(out)
```

# apprentissage des erreurs

dada2 calcul 1 modèle d'erreur a partir des données de sequencage on applique cette méthode sur les Readfor 1 et les reverses

```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
```
```{r}
errR <- learnErrors(filtRs, multithread=TRUE)
```
```{r}
plotErrors(errF, nominalQ=TRUE)
```
```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```

```{r}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```
```{r}
dadaFs[[1]]
```
# aligner les R1 et les R2 en 1 contig 

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```


# maintenant on va construire la table d'observation 

```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```
```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```
# éliminer les sequences chimériques 
# remaque on ne travail ici plus que sur les contig, pas les Reads

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```
```{r}
1-sum(seqtab.nochim)/sum(seqtab)
```
il y a 3,5% de données chimériques dnas notre jeu de données

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```

```{bash}
wget https://zenodo.org/record/3986799/files/silva_nr99_v138_train_set.fa.gz
```

```{r}
taxa <- assignTaxonomy(seqtab.nochim, "~/silva_nr99_v138_train_set.fa.gz", multithread=TRUE)
```

# assignement taxonomique 

```{bash}
wget https://zenodo.org/record/3986799/files/silva_species_assignment_v138.fa.gz
```

```{r}
taxa <- addSpecies(taxa, "~/silva_species_assignment_v138.fa.gz")
```

```{r}
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

# Evaluate accuracy

```{r}
unqs.mock <- seqtab.nochim["Mock",]
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
```
```{r}
mock.ref <- getSequences(file.path(path, "HMP_MOCK.v35.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```



